---
title: "Hands-on 3 - Processing and Visualising Flow DataR"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
  
highlight-style: github
code:
  style: "color: red;"
---

## 15.1 Overview 

Spatial interaction represent the flow of people, material, or information between locations in geographical space. It encompasses everything from freight shipments, energy flows, and the global trade in rare antiquities, to flight schedules, rush hour woes, and pedestrian foot traffic.

Each spatial interaction, as an analogy for a set of movements, is composed of a discrete origin/destination pair. Each pair can be represented as a cell in a matrix where rows are related to the locations (centroids) of origin, while columns are related to locations (centroids) of destination. Such a matrix is commonly known as an origin/destination matrix, or a spatial interaction matrix.

In this hands-on exercise, you will learn how to build an OD matrix by using Passenger Volume by Origin Destination Bus Stops data set downloaded from LTA DataMall. By the end of this hands-on exercise, you will be able:

-   to import and extract OD data for a selected time interval, 

-   to import and save geospatial data (i.e. bus stops and mpsz) into sf tibble data frame objects, 

-   to populate planning subzone code into bus stops sf tibble data frame, 

-   to construct desire lines geospatial data from the OD data, and 

-   to visualise passenger volume by origin and destination bus stops by using the desire lines data. 

## 15.2 Getting Started 

For the purpose of this exercise, four r packages will be used. They are:

sf for importing, integrating, processing and transforming geospatial data. tidyverse for importing, integrating, wrangling and visualising data. tmap for creating thematic maps.

```{r}
pacman::p_load(tmap, sf, DT, performance, ggpubr, stplanr, tidyverse)
```


## 15.3 

Preparing the Flow Data 15.3.1 Importing the OD data Firstly, we will import the Passenger Volume by Origin Destination Bus Stops data set downloaded from LTA DataMall by using read_csv() of readr package.


```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202310.csv")
```


Let use display the odbus tibble data table by using the code chunk below.

```{r}
glimpse(odbus)
```


```{r}

odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```


### 15.3.2 Extracting the study data 

For the purpose of this exercise, we will extract commuting flows on weekday and between 6 and 9 o'clock.


```{r}
odbus6_9 <- odbus %>% 
  filter(DAY_TYPE == "WEEKDAY") %>% 
  filter(TIME_PER_HOUR >= 6 & TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE, DESTINATION_PT_CODE) %>% 
  summarise(TRIPS = sum(TOTAL_TRIPS))

write_rds(odbus6_9,"data/rds/odbus6_9.rds")

odbus6_9 <- read_rds("data/rds/odbus6_9.rds")
```

Table below shows the content of odbus6_9

```{r}
datatable(odbus6_9)
```

## 15.4 Working with Geospatial Data 

For the purpose of this exercise, two geospatial data will be used. They are:

-   BusStop: This data provides the location of bus stop as at last quarter of 2022. 

-   MPSZ-2019: This data provides the sub-zone boundary of URA Master Plan 2019. Both data sets are in ESRI shapefile format.

### 15.4.1 Importing geospatial data 

Two geospatial data will be used in this exercise, they are:

```{r}
busstop <- st_read(dsn =  "data/geospatial", layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- st_read(dsn =  "data/geospatial", layer = "MPSZ-2019") %>% 
  st_transform(crs = 3414)
```

Note st_read() function of sf package is used to import the shapefile into R as sf data frame. st_transform() function of sf package is used to transform the projection to crs 3414. 

## 15.5 Geospatial data wrangling 

### 15.5.1 Combining Busstop and mpsz 

Code chunk below populates the planning subzone code (i.e. SUBZONE_C) of mpsz sf data frame into busstop sf data frame.

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>% 
  select(BUS_STOP_N, SUBZONE_C) %>% 
  st_drop_geometry()

write_rds(busstop_mpsz,  "data/rds/busstop_mpsz.rds")
```

Next, we are going to append the planning subzone code from busstop_mpsz data frame onto odbus6_9 data frame.

```{r}
od_data <- left_join(odbus6_9 , busstop_mpsz, by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>% 
  rename(ORIGIN_BS = ORIGIN_PT_CODE, ORIGIN_SZ = SUBZONE_C, DESTIN_BS = DESTINATION_PT_CODE)
```

Before continue, it is a good practice for us to check for duplicating records.

```{r}
duplicate <- od_data %>% 
  group_by_all() %>% 
  filter(n()>1) %>%
  ungroup()
duplicate
```

If duplicated records are found, the code chunk below will be used to retain the unique records.

```{r}
od_data <- unique(od_data)
```

It will be a good practice to confirm if the duplicating records issue has been addressed fully.

Next, we will update od_data data frame cwith the planning subzone codes.

```{r}

od_data <- left_join(od_data , busstop_mpsz, by = c("DESTIN_BS" = "BUS_STOP_N"))

duplicate <- od_data %>% 
  group_by_all() %>%
  filter(n()>1) %>% 
  ungroup()
duplicate

```



```{r}
od_data <- unique(od_data)

od_data <- od_data %>% 
rename(DESTIN_SZ = SUBZONE_C) %>% 
  drop_na() %>% 
  group_by(ORIGIN_SZ, DESTIN_SZ) %>% 
  summarise(MORNING_PEAK = sum(TRIPS))


write_rds(od_data,  "data/rds/od_data.rds")

od_data <- read_rds( "data/rds/od_data.rds")

```

## 15.6 Visualising Spatial Interaction 

In this section, you will learn how to prepare a desire line by using stplanr package.

### 15.6.1 Removing intra-zonal flows 

We will not plot the intra-zonal flows. The code chunk below will be used to remove intra-zonal flows.


```{r}
od_data1 <- od_data[od_data$ORIGIN_SZ!=od_data$DESTIN_SZ,]
```

### 15.6.2 Creating desire lines 

In this code chunk below, od2line() of stplanr package is used to create the desire lines.

```{r}
flowLine <- od2line(flow = od_data1, zones = mpsz, zone_code = "SUBZONE_C")
```

### 15.6.3 Visualising the desire lines 

To visualise the resulting desire lines, the code chunk below is used.

```{r}
tm_shape(mpsz) + tm_polygons() + flowLine %>%
  tm_shape() + tm_lines(lwd = "MORNING_PEAK", style = "quantile", scale = c(0.1, 1, 3, 5, 7, 10), n = 6, alpha = 0.3)
```

When the flow data are very messy and highly skewed like the one shown above, it is wiser to focus on selected flows, for example flow greater than or equal to 5000 as shown below.

```{r}
tm_shape(mpsz) + tm_polygons() + flowLine %>%
filter(MORNING_PEAK >= 5000) %>% 
  tm_shape() + 
  tm_lines(lwd = "MORNING_PEAK", 
           style = "quantile", 
           scale = c(0.1, 1, 3, 5, 7, 10), 
           n = 6, 
           alpha = 0.3) 
```

```
