---
title: "In-class Exercise 3 - Calibrating Spatial Interaction Models with R"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
  
highlight-style: github
code:
  style: "color: red;"
---

## Overview

This in-class introduces an alternative R package to spdep package you used in Hands-on Exercise 6. The package is called [**sfdep**](https://sfdep.josiahparry.com/index.html). According to Josiah Parry, the developer of the package, "sfdep builds on the great shoulders of **spdep** package for spatial dependence. sfdep creates an sf and tidyverse friendly interface to the package as well as introduces new functionality that is not present in spdep. sfdep utilizes list columns extensively to make this interface possible."

## Getting started

```{r}
pacman::p_load(tmap, sf, sp, DT,
               performance, reshape2,
               ggpubr, units, tidyverse)
```

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                   layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)

mpsz_sp <- as(mpsz, "Spatial")
mpsz_sp
```

```{r}
dist <- spDists(mpsz_sp, 
                longlat = FALSE)
head(dist, n=c(10, 10))



```

### 16.5.3 Labelling column and row heanders of a distance matrix

```{r}
sz_names <- mpsz$SUBZONE_C
colnames(dist) <- paste0(sz_names)
rownames(dist) <- paste0(sz_names)


```

```{r}
distPair <- melt(dist) %>%
  rename(dist = value)
head(distPair, 10)



```

### 16.5.5 Updating intra-zonal distances

In this section, we are going to append a constant value to replace the intra-zonal distance of 0.

First, we will select and find out the minimum value of the distance by using summary().

```{r}
distPair %>%
  filter(dist > 0) %>%
  summary()
```

Next, a constant distance value of 50m is added into intra-zones distance. \> 50 is derived from approximately minimum of 173.8 (found out earlier in summary statistics) divided by 2. Note : \*Intra-zone

```{r}
distPair$dist <- ifelse(distPair$dist == 0,
                        50, distPair$dist)
distPair %>%
  summary()
```

```{r}
distPair <- distPair %>%
  rename(orig = Var1,
         dest = Var2)

write_rds(distPair, "data/rds/distPair.rds") 
```

## 16.6 Preparing flow data

```{r}
od_data <- read_rds("data/rds/od_data.rds")

flow_data <- od_data %>%
  group_by(ORIGIN_SZ, DESTIN_SZ) %>% 
  summarize(TRIPS = sum(MORNING_PEAK)) 

head(flow_data, 10)
```

### 16.6.1 Separating intra-flow from passenger volume df

```{r}
flow_data$FlowNoIntra <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ, 
  0, flow_data$TRIPS)
flow_data$offset <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ, 
  0.000001, 1)
```

### 16.6.2 Combining passenger volume data with distance value

Before we can join flow_data and distPair, we need to convert data value type of ORIGIN_SZ and DESTIN_SZ fields of flow_data dataframe into factor data type.

```{r}
flow_data$ORIGIN_SZ <- as.factor(flow_data$ORIGIN_SZ)
flow_data$DESTIN_SZ <- as.factor(flow_data$DESTIN_SZ)
```

Now, left_join() of dplyr will be used to merge flow_data dataframe and distPair dataframe. The output is called flow_data1.

```{r}
flow_data1 <- flow_data %>%
  left_join (distPair,
             by = c("ORIGIN_SZ" = "orig",
                    "DESTIN_SZ" = "dest"))

```

## 16.7 Preparing Origin and Destination Attributes

### 16.7.1 Importing population data

```{r}
pop <- read_csv("data/aspatial/pop.csv")
```

### 16.7.2 Geospatial data wrangling

```{r}
pop <- pop %>%
  left_join(mpsz,
            by = c("PA" = "PLN_AREA_N",
                   "SZ" = "SUBZONE_N")) %>%
  select(1:6, everything()) %>%
  rename(SZ_NAME = SZ,
         SZ = SUBZONE_C)

```

### 16.7.3 Preparing origin attribute

```{r}
flow_data1 <- flow_data1 %>%
  left_join(pop,
            by = c(ORIGIN_SZ = "SZ")) %>%
  rename(ORIGIN_AGE7_12 = AGE7_12,
         ORIGIN_AGE13_24 = AGE13_24,
         ORIGIN_AGE25_64 = AGE25_64) %>%
  select(-c(PA, SZ_NAME))

```

### 16.7.4 Preparing destination attribute

```{r}
flow_data1 <- flow_data1 %>%
  left_join(pop,
            by = c(DESTIN_SZ = "SZ")) %>%
  select(-c(PA, SZ_NAME)) %>%
  rename(DESTIN_AGE7_12 = AGE7_12,
         DESTIN_AGE13_24 = AGE13_24,
         DESTIN_AGE25_64 = AGE25_64) 
  
write_rds(flow_data1, "data/rds/SIM_data")
```

## **16.8 Calibrating Spatial Interaction Models**

In this section, you will learn how to calibrate Spatial Interaction Models by using Poisson Regression method.

### **16.8.1 Importing the modelling data**

Firstly, let us import the modelling data by using the code chunk below.

```{r}
SIM_data <- read_rds("data/rds/SIM_data.rds")

```

### **16.8.2 Visualising the dependent variable**

Firstly, let us plot the distribution of the dependent variable (i.e. TRIPS) by using histogram method by using the code chunk below.

```{r}
ggplot(data = SIM_data,
       aes(x = TRIPS)) +
  geom_histogram()
```

Notice that the distribution is highly skewed and not resemble bell shape or also known as normal distribution.

Next, let us visualise the relation between the dependent variable and one of the key independent variable in Spatial Interaction Model, namely distance.

```{r}
ggplot(data = SIM_data,
       aes(x = dist,
           y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)
```

Notice that their relationship hardly resemble linear relationship.

On the other hand, if we plot the scatter plot by using the log transformed version of both variables, we can see that their relationship is more resemble linear relationship.

```{r}
ggplot(data = SIM_data,
       aes(x = log(dist),
           y = log(TRIPS))) +
  geom_point() +
  geom_smooth(method = lm)
```

### **16.8.3 Checking for variables with zero values**

Since Poisson Regression is based of log and log 0 is undefined, it is important for us to ensure that no 0 values in the explanatory variables.

In the code chunk below, summary() of Base R is used to compute the summary statistics of all variables in *SIM_data* data frame.

```{r}
summary(SIM_data)
```

The print report above reveals that variables ORIGIN_AGE7_12, ORIGIN_AGE13_24, ORIGIN_AGE25_64,DESTIN_AGE7_12, DESTIN_AGE13_24, DESTIN_AGE25_64 consist of 0 values.

In view of this, code chunk below will be used to replace zero values to 0.99.

```{r}
SIM_data$DESTIN_AGE7_12 <- ifelse(
  SIM_data$DESTIN_AGE7_12 == 0,
  0.99, SIM_data$DESTIN_AGE7_12)
SIM_data$DESTIN_AGE13_24 <- ifelse(
  SIM_data$DESTIN_AGE13_24 == 0,
  0.99, SIM_data$DESTIN_AGE13_24)
SIM_data$DESTIN_AGE25_64 <- ifelse(
  SIM_data$DESTIN_AGE25_64 == 0,
  0.99, SIM_data$DESTIN_AGE25_64)
SIM_data$ORIGIN_AGE7_12 <- ifelse(
  SIM_data$ORIGIN_AGE7_12 == 0,
  0.99, SIM_data$ORIGIN_AGE7_12)
SIM_data$ORIGIN_AGE13_24 <- ifelse(
  SIM_data$ORIGIN_AGE13_24 == 0,
  0.99, SIM_data$ORIGIN_AGE13_24)
SIM_data$ORIGIN_AGE25_64 <- ifelse(
  SIM_data$ORIGIN_AGE25_64 == 0,
  0.99, SIM_data$ORIGIN_AGE25_64)
```

Notice that all the 0 values have been replaced by 0.99.

### **16.8.4 Unconstrained Spatial Interaction Model**

In this section, you will learn how to calibrate an unconstrained spatial interaction model by using `glm()` of Base Stats. The explanatory variables are origin population by different age cohort, destination population by different age cohort (i.e. *ORIGIN_AGE25_64*) and distance between origin and destination in km (i.e. *dist*).

The general formula of Unconstrained Spatial Interaction Model

```{r}
uncSIM <- glm(formula = TRIPS ~ 
                log(ORIGIN_AGE25_64) + 
                log(DESTIN_AGE25_64) +
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
uncSIM

```

### **16.8.5 R-squared function**

In order to measure how much variation of the trips can be accounted by the model we will write a function to calculate R-Squared value as shown below.

```{r}
CalcRSquared <- function(observed,estimated){
  r <- cor(observed,estimated)
  R2 <- r^2
  R2
}
```

Next, we will compute the R-squared of the unconstrained SIM by using the code chunk below.

```{r}
CalcRSquared(uncSIM$data$TRIPS, uncSIM$fitted.values)

```

```{r}
r2_mcfadden(uncSIM)

```

### **16.8.6 Origin (Production) constrained SIM**

In this section, we will fit an origin constrained SIM by using the code3 chunk below.

The general formula of Origin Constrained Spatial Interaction Model

```{r}
orcSIM <- glm(formula = TRIPS ~ 
                 ORIGIN_SZ +
                 log(DESTIN_AGE25_64) +
                 log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)

```

We can examine how the constraints hold for destinations this time.

```{r}
CalcRSquared(orcSIM$data$TRIPS, orcSIM$fitted.values)

```

### 16.8.7 Destination constrained

In this section, we will fit a destination constrained SIM by using the code chunk below.

```{r}
decSIM <- glm(formula = TRIPS ~ 
                DESTIN_SZ + 
                log(ORIGIN_AGE25_64) + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
```

We can examine how the constraints hold for destinations this time.

```{r}
CalcRSquared(decSIM$data$TRIPS, decSIM$fitted.values)
```

### 16.8.8 Doubly constrained

In this section, we will fit a doubly constrained SIM by using the code chunk below.

```{r}
dbcSIM <- glm(formula = TRIPS ~ 
                ORIGIN_SZ + 
                DESTIN_SZ + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
```

We can examine how the constraints hold for destinations this time.

```{r}
CalcRSquared(dbcSIM$data$TRIPS, dbcSIM$fitted.values)
```

Notice that there is a relatively greater improvement in the R\^2 value.

### **16.8.9 Model comparison**

Another useful model performance measure for continuous dependent variable is [Root Mean Squared Error](https://towardsdatascience.com/what-does-rmse-really-mean-806b65f2e48e). In this sub-section, you will learn how to use [`compare_performance()`](https://easystats.github.io/performance/reference/compare_performance.html) of [**performance**](https://easystats.github.io/performance/) package

First of all, let us create a list called *model_list* by using the code chunk below.

```{r}
model_list <- list(unconstrained=uncSIM,
                   originConstrained=orcSIM,
                   destinationConstrained=decSIM,
                   doublyConstrained=dbcSIM)
```

Next, we will compute the RMSE of all the models in *model_list* file by using the code chunk below.

```{r}
compare_performance(model_list,
                    metrics = "RMSE")
```

The print above reveals that doubly constrained SIM is the best model among all the four SIMs because it has the smallest RMSE value of 1487.111.

### **16.8.10 Visualising fitted**

In this section, you will learn how to visualise the observed values and the fitted values.

Firstly we will extract the fitted values from each model by using the code chunk below.

```{r}
df <- as.data.frame(uncSIM$fitted.values) %>%
  round(digits = 0)
```

Next, we will join the values to *SIM_data* data frame.

```{r}
SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(uncTRIPS = "uncSIM$fitted.values")
```

Repeat the same step by for Origin Constrained SIM (i.e. orcSIM)

```{r}
df <- as.data.frame(orcSIM$fitted.values) %>%
  round(digits = 0)

SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(orcTRIPS = "orcSIM$fitted.values")
```

Repeat the same step by for Destination Constrained SIM (i.e. decSIM)

```{r}
df <- as.data.frame(decSIM$fitted.values) %>%
  round(digits = 0)
```

Repeat the same step by for Doubly Constrained SIM (i.e. dbcSIM)

```{r}
df <- as.data.frame(dbcSIM$fitted.values) %>%
  round(digits = 0)

SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(dbcTRIPS = "dbcSIM$fitted.values")
unc_p <- ggplot(data = SIM_data,
                aes(x = uncTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

orc_p <- ggplot(data = SIM_data,
                aes(x = orcTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)


dec_p <- ggplot(data = SIM_data,
                aes(x = decTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

dbc_p <- ggplot(data = SIM_data,
                aes(x = dbcTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

```
